# ============================================================
# LOCK HOSPITAL DIGITAL HISTORY PROJECT
# Data Cleanup and Standardization 
# Author: Megha Khanna
# ============================================================

```{r}
# all the libraries I installed for this process
library(DBI)
library(RSQLite)
library(readxl)
library(dplyr)
library(stringr)
library(shiny)
library(plotly)
library(tidytext)
library(tidyr)

```
I installed DBI, which is required for interfacing R with databases, and RSQLite to enable interaction with SQLite files. Since my datasets were quite large, I transitioned from Excel to SQLite for more efficient querying and storage. To make my project interactive and user-friendly, I chose the Shiny framework and used Plotly to build interactive visualizations.
```{r}

# --- Load Required Libraries ---
library(DBI)
library(RSQLite)
library(readxl)
library(dplyr)
library(stringr)

excel_path  <- "data_raw/DS_Dataset.xlsx"
sqlite_path <- "data_processed/primary_sources.sqlite"

```

```{r}
# --- Connect to SQLite Database ---
con <- dbConnect(RSQLite::SQLite(), sqlite_path)
# Source: https://cran.r-project.org/web/packages/RSQLite/vignettes/RSQLite.html

# --- Import Excel Sheets and Write to SQLite Tables ---
sheets <- excel_sheets(excel_path)
for (sheet in sheets) {
  data <- read_excel(excel_path, sheet = sheet)
  dbWriteTable(con, sheet, data, overwrite = TRUE)
}
# Source: https://www.rdocumentation.org/packages/DBI/versions/1.2.2/topics/dbWriteTable

# --- Verify Tables Created ---
dbListTables(con)

```
```{r}
# --- Standardize Legislative Act Names ---
# Purpose: Harmonize historical references to the Contagious Diseases Acts
#          and Cantonment Acts for consistency across reports.
# Sources:
# - DBI::dbExecute(): https://www.rdocumentation.org/packages/DBI/versions/1.2.2/topics/dbExecute

replacements <- list(
  "Act XIV/1868"                       = "Act XIV of 1868",
  "Contagious Diseases Act XIV (1868)" = "Act XIV of 1868",
  "Act XXII/1864"                      = "Act XXII of 1864",
  "Cantonment Act XXII (1864)"         = "Act XXII of 1864"
)

for (old_value in names(replacements)) {
  new_value <- replacements[[old_value]]
  query <- sprintf(
    "UPDATE Hospitals
     SET act = '%s'
     WHERE act = '%s';",
    new_value, old_value
  )
  dbExecute(con, query)
}

# Verify updates worked
dbGetQuery(con, "SELECT DISTINCT act FROM Hospitals LIMIT 5")

```
```{r}
# --- Standardize Hospital Classifications ---
# Purpose: Normalize inconsistent text values (e.g., "1st Class") into uniform categories.

dbExecute(con, "UPDATE Hospitals SET class = 'First Class'  WHERE class IN ('1st Class')")
dbExecute(con, "UPDATE Hospitals SET class = 'Second Class' WHERE class IN ('2nd Class','First/Second Class')")
dbExecute(con, "UPDATE Hospitals SET class = 'Third Class'  WHERE class IN ('3rd Class','Third Class (formerly Second)')")
dbExecute(con, "UPDATE Hospitals SET class = 'Second Class' WHERE class LIKE 'First Class to 1 Sept%'")

```
```{r}
# --- Standardize Country and Region Names ---
# Purpose: Consolidate colonial-era territorial references for historical accuracy.
# Sources: https://www.rdocumentation.org/packages/DBI/versions/1.2.2/topics/dbExecute

# Country-level harmonization
dbExecute(con, "UPDATE women_admission  SET country = 'British India' WHERE country = 'India'")
dbExecute(con, "UPDATE troops_admission SET country = 'British India' WHERE country = 'India'")
dbExecute(con, "UPDATE women_admission  SET country = 'British Burma' WHERE country = 'Burma, India (British Burma)'")
dbExecute(con, "UPDATE troops_admission SET country = 'British Burma' WHERE country = 'Burma, India (British Burma)'")

# Region-level harmonization
dbExecute(con, "UPDATE women_admission  SET region = 'Madras Presidency' WHERE region = 'Madras Region'")
dbExecute(con, "UPDATE troops_admission SET region = 'Madras Presidency' WHERE region = 'Madras Region'")
dbExecute(con, "UPDATE women_admission  SET region = 'British Burma Division' WHERE region IN ('Burma','British Burma')")
dbExecute(con, "UPDATE troops_admission SET region = 'British Burma Division' WHERE region IN ('Burma','British Burma')")

```
```{r}
# --- Clean and Convert Staff Data Columns ---
# Purpose: Extract numeric counts from descriptive staff fields for quantitative analysis.
# Sources:
# - dplyr::mutate(): https://www.rdocumentation.org/packages/dplyr/versions/1.1.4/topics/mutate
# - stringr::str_extract(): https://www.rdocumentation.org/packages/stringr/versions/1.5.1/topics/str_extract

# Read the Hospitals table
hospitals <- dbReadTable(con, "Hospitals")

# Extract numbers and convert to integers
hospitals_cleaned <- hospitals %>%
  mutate(
    staff_medical_officers    = as.integer(str_extract(staff_medical_officers, "\\d+")),
    staff_hospital_assistants = as.integer(str_extract(staff_hospital_assistants, "\\d+")),
    staff_matron              = as.integer(str_extract(staff_matron, "\\d+")),
    staff_coolies             = as.integer(str_extract(staff_coolies, "\\d+")),
    staff_peons               = as.integer(str_extract(staff_peons, "\\d+")),
    staff_watermen            = as.integer(str_extract(staff_watermen, "\\d+"))
  )

# Write the cleaned version back into the database
dbWriteTable(con, "Hospitals", hospitals_cleaned, overwrite = TRUE)

# Final small regional fix
dbExecute(con, "
  UPDATE troops_admission
  SET region = 'Madras Presidency'
  WHERE region = 'Madras Region'
")

```
```{r}
# --- Standardize Country and Region Names ---
# Purpose: Standardize region and country names for all future datasets.
# Source: Inspired by CRAN’s dplyr::case_when() and string standardization workflows
# https://www.rdocumentation.org/packages/dplyr/versions/1.1.4/topics/case_when

standardize_rc <- function(df) {
  df %>%
    mutate(
      region = case_when(
        region %in% c("Burma", "Burma (Madras Presidency Civil)", "India (British Burma)") ~ "British Burma",
        region %in% c("Madras Presidency", "Madras Presidency (Civil)") ~ "Madras Presidency",
        TRUE ~ region
      ),
      country = case_when(
        country %in% c("India", "British India; India", "British India", "India (British)") ~ "British India",
        country %in% c("Burma", "India (British Burma)") ~ "British Burma",
        TRUE ~ country
      )
    )
}

```

